---
title: "Confidence Interval and Confidence Band — Survival Tutorial (RLT)"
author: "Ruoqing Zhu"
date: "Last Updated: `r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    code_folding: hide
    df_print: paged
    toc: yes
    toc_float:
      collapsed: true
      smooth_scroll: true 
    toc_depth: '2'
  pdf_document:
    toc: yes
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE, warning = FALSE,
  fig.width = 9, fig.height = 7, out.width = "90%", fig.align = "center",
  collapse = TRUE, class.source = "fold-show"
)
```

```{r load-libs}
library(RLT)
library(ggplot2)
```

## Overview

This page demonstrates how to construct **pointwise confidence intervals** and **simultaneous confidence bands** for individual survival curves predicted by RLT. The bands use a smoothed low-rank covariance approach — GAM-smoothed eigenvalue decomposition plus an eigenvalue-ratio weighted residual correction — which produces stable, well-calibrated bands even when the number of time points is large relative to the number of trees.

## Data

We simulate right-censored survival data from a proportional hazards model with exponential event times. The first and third predictors carry signal; the rest are noise.

```{r data-prep}
set.seed(2)

n <- 200
p <- 10
X <- matrix(rnorm(n * p), n, p)

xlink <- function(x) exp(x[, 1] + x[, 3] / 2)
FT <- rexp(n, rate = xlink(X))
CT <- pmin(6, rexp(n, rate = 0.25))

Y <- pmin(FT, CT)
Censor <- as.numeric(FT <= CT)

# Test subjects for visualization
ntest  <- 4
set.seed(100)
testX  <- matrix(rnorm(ntest * p), ntest, p)

# True survival for reference
timepoints <- sort(unique(Y[Censor == 1]))
SurvMat    <- matrix(NA, nrow(testX), length(timepoints))
exprate    <- xlink(testX)
for (j in seq_along(timepoints)) {
  SurvMat[, j] <- 1 - pexp(timepoints[j], rate = exprate)
}
```

## Fit and predict with variance

We use `var.mode = "matched"` (matched-sample U-statistic) for variance estimation. This automatically adjusts sampling to subsampling without replacement at 50%.

```{r fit-predict}
fit <- RLT(
  X, Y, Censor, model = "survival",
  ntrees = 1000, mtry = min(p, 10), nmin = 5,
  split.gen = "random", nsplit = 3,
  resample.prob = 0.8, resample.replace = FALSE,
  importance = FALSE, verbose = FALSE,
  ncores = 1,
  var.mode = "matched",
  param.control = list(split.rule = "logrank")
)

# Predict survival curves and covariance for test subjects
RLTPred <- predict(fit, testX, var.est = TRUE, ncores = 1)
```

## Pointwise confidence intervals

Before constructing simultaneous bands, it is useful to look at pointwise intervals ($\pm 1.96 \times \text{SD}$ from the diagonal of the covariance matrix). These are **not** simultaneous — each time point is considered independently.

```{r plot-helper}
# Helper: build a data.frame for ggplot from one subject
build_subject_df <- function(subject_id, pred, SurvMat = NULL, band_list = NULL) {
  tp  <- pred$timepoints
  S   <- as.numeric(pred$Survival[subject_id, ])
  sdf <- data.frame(
    time  = tp,
    surv  = pmin(pmax(S, 0), 1),
    lower_pw = pmin(pmax(S - qnorm(0.975) * sqrt(diag(pred$Cov[,, subject_id])), 0), 1),
    upper_pw = pmin(pmax(S + qnorm(0.975) * sqrt(diag(pred$Cov[,, subject_id])), 0), 1)
  )
  
  if (!is.null(SurvMat)) {
    sdf$truth <- pmin(pmax(as.numeric(SurvMat[subject_id, ]), 0), 1)
  }
  
  if (!is.null(band_list)) {
    sdf$lower_band <- pmin(pmax(as.numeric(band_list[[subject_id]]$lower), 0), 1)
    sdf$upper_band <- pmin(pmax(as.numeric(band_list[[subject_id]]$upper), 0), 1)
  }
  
  sdf
}
```

```{r pointwise-plot, fig.width=10, fig.height=9}
# Build data for all subjects
plot_data <- do.call(rbind, lapply(1:ntest, function(i) {
  d <- build_subject_df(i, RLTPred, SurvMat)
  d$subject <- paste("Subject", i)
  d
}))

ggplot(plot_data, aes(x = time)) +
  facet_wrap(~ subject, nrow = 2, scales = "free_y") +
  # Pointwise band (shaded)
  geom_ribbon(aes(ymin = lower_pw, ymax = upper_pw),
              fill = "grey80", alpha = 0.6) +
  # Truth
  geom_line(aes(y = truth, colour = "True S(t)"), linewidth = 0.9, linetype = "dashed") +
  # Estimate
  geom_step(aes(y = surv, colour = "Estimated S(t)"), linewidth = 0.9) +
  scale_colour_manual(name = NULL,
                      values = c("True S(t)" = "#E41A1C", "Estimated S(t)" = "#222222")) +
  scale_y_continuous(limits = c(0, 1), expand = c(0.01, 0.01)) +
  labs(x = "Time", y = "Survival Probability",
       title = "Pointwise 95% Confidence Intervals") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom",
        panel.grid.minor = element_blank(),
        strip.text = element_text(face = "bold"))
```

## Simultaneous confidence band

`get.surv.band()` with `approach = "smoothed"` constructs a simultaneous confidence band using a GAM-smoothed low-rank approximation of the covariance matrix, plus an eigenvalue-ratio weighted residual correction. The smoothed approach is more stable than the naive Monte Carlo method, especially when the number of time points is large.

```{r band-calc}
SurvBand <- get.surv.band(RLTPred, alpha = 0.05, approach = "smoothed", k_rank = 10)
```

```{r band-plot, fig.width=10, fig.height=9}
plot_band <- do.call(rbind, lapply(1:ntest, function(i) {
  d <- build_subject_df(i, RLTPred, SurvMat, SurvBand)
  d$subject <- paste("Subject", i)
  d
}))

ggplot(plot_band, aes(x = time)) +
  facet_wrap(~ subject, nrow = 2, scales = "free_y") +
  # Simultaneous band (blue shading)
  geom_ribbon(aes(ymin = lower_band, ymax = upper_band),
              fill = "#3B82F6", alpha = 0.25) +
  # Pointwise band (grey shading, for comparison)
  geom_ribbon(aes(ymin = lower_pw, ymax = upper_pw),
              fill = "grey70", alpha = 0.3) +
  # Truth
  geom_line(aes(y = truth, colour = "True S(t)"), linewidth = 0.9, linetype = "dashed") +
  # Estimate
  geom_step(aes(y = surv, colour = "Estimated S(t)"), linewidth = 0.9) +
  scale_colour_manual(name = NULL,
                      values = c("True S(t)" = "#E41A1C", "Estimated S(t)" = "#222222")) +
  scale_y_continuous(limits = c(0, 1), expand = c(0.01, 0.01)) +
  labs(x = "Time", y = "Survival Probability",
       title = "Smoothed Covariance — 95% Simultaneous Confidence Band",
       subtitle = "Grey = pointwise CI; Blue = simultaneous band") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom",
        panel.grid.minor = element_blank(),
        strip.text = element_text(face = "bold"))
```

## (Optional) Proportion-selected smoothing

The smoothed approach can also choose the rank by cumulative eigenvalue proportion instead of a fixed `k_rank`:

```r
SurvBand_prop <- get.surv.band(
  RLTPred, alpha = 0.05,
  approach = "smoothed",
  k_mode = "proportion",
  k_prop = 0.95
)
```

You can increase `nsim` for stability at the cost of runtime.

## (Optional) Reducing the time grid

For large datasets, the full set of failure times can make covariance matrices unwieldy. Use `band.grid.size` in `predict()` to evaluate variance on a reduced quantile-based grid:

```r
pred_reduced <- predict(fit, testX, var.est = TRUE, band.grid.size = 50)
length(pred_reduced$timepoints)  # <= 50 time points
```
