---
title: "RLT with Linear Combinations"
author: "Ruoqing Zhu"
date: "Last Updated: `r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    code_folding: hide
    df_print: paged
    toc: yes
    toc_float:
      collapsed: true
      smooth_scroll: true
    toc_depth: '2'
  pdf_document:
    toc: yes
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 5, out.width = "70%", fig.align = "center",
  collapse = TRUE, class.source = "fold-show"
)
```

## Overview

Linear-combination splits allow each internal node to split on a weighted combination of several variables rather than a single predictor. This can help when the signal is carried by a direction such as `X1 + X2` instead of one coordinate at a time.

The example below uses regression. The same idea can also be applied to classification and survival forests by setting `model = "classification"` or `model = "survival"` and choosing a model-appropriate `linear.comb.method`.

## Regression Example

```{r regression-data}
set.seed(3)

n <- 160
p <- 6
X <- matrix(rnorm(n * p), n, p)
colnames(X) <- paste0("X", seq_len(p))

y <- 1 + X[, 1] + X[, 2] + 0.5 * X[, 3] + rnorm(n, sd = 0.5)

train_id <- 1:120
trainX <- X[train_id, ]
trainY <- y[train_id]
testX <- X[-train_id, ]
testY <- y[-train_id]
```

Fit a regression forest that considers three-variable linear combinations at a split.

```{r linear-combination-fit}
library(RLT)

fit_lc <- RLT(
  trainX, trainY,
  model = "regression",
  ntrees = 200,
  nmin = 5,
  mtry = 3,
  split.gen = "random",
  nsplit = 3,
  linear.comb = 3,
  linear.comb.method = "sir",
  ncores = 1,
  verbose = FALSE
)

print(fit_lc)
```

## Prediction

```{r linear-combination-predict}
pred <- predict(fit_lc, testX, ncores = 1)
test_pred <- if (is.list(pred) && !is.null(pred$Prediction)) {
  pred$Prediction
} else {
  as.numeric(pred)
}

list(
  Test_MSE = round(mean((test_pred - testY)^2), 4),
  Linear_Combination_Size = fit_lc$parameters$linear.comb
)
```

## Other Model Types

For classification, use `model = "classification"` with methods such as `"lda"`, `"naive"`, `"random"`, or `"logistic"`.

For survival, use `model = "survival"` with methods such as `"coxph"` or `"naive"`.
