---
title: "Variable Importance (Permutation and Distributed) - Tutorial (RLT)"
author: "Ruoqing Zhu"
date: "Last Updated: `r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    code_folding: hide
    df_print: paged
    toc: yes
    toc_float:
      collapsed: true
      smooth_scroll: true 
    toc_depth: '2'
  pdf_document:
    toc: yes
    toc_depth: 2
---

```{r setup, include=FALSE}
# Clean, consistent output for all chunks
knitr::opts_chunk$set(
  message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 5, out.width = "80%", fig.align = "center",
  collapse = TRUE, class.source = "fold-show"
)
```

## Overview

This page demonstrates how to obtain and visualize variable importance (VI) from RLT.

## Data

We generate continuous and categorical predictors with a continuous outcome.

```{r data-prep}
# (Optional) For reproducibility in this tutorial only.
set.seed(1)

# ---- Generate a small synthetic dataset ----
trainn <- 80
testn  <- 20
n <- trainn + testn
p <- 10

# Continuous + categorical predictors (last half as factors)
X1 <- matrix(rnorm(n * (p/2)), n, p/2)
X2 <- matrix(as.integer(runif(n * (p/2)) * 3), n, p/2)  # integers 0,1,2

# Continuous outcome with a simple signal + noise
X_numeric <- data.frame(X1, X2)
y <- 1 + rowSums(X_numeric[, 2:6]) +
  2 * (X_numeric[, p/2 + 1] %in% c(1, 2)) + rnorm(n)

X <- X_numeric
X[, (p/2 + 1):p] <- lapply(X[, (p/2 + 1):p], as.factor)

# Train / test split
trainX <- X[1:trainn, ]
trainY <- y[1:trainn]
testX  <- X[(trainn + 1):(trainn + testn), ]
testY  <- y[(trainn + 1):(trainn + testn)]
```

## Option A - Permutation Importance

Set `importance` to enable permutation-based VI during training.

```{r perm-importance}
# install.packages("devtools"); devtools::install_github("teazrq/RLT")
library(RLT)

# Minimal, sensible defaults
ntrees <- 200
ncores <- 1
nmin   <- 5
mtry   <- p/2
samplereplace <- TRUE
sampleprob    <- 0.80
rule    <- "best"
nsplit  <- ifelse(rule == "best", 0, 3)

fit_perm <- RLT(
  trainX, trainY, model = "regression",
  ntrees = ntrees, mtry = mtry, nmin = nmin,
  resample.prob = sampleprob, split.gen = rule,
  resample.replace = samplereplace,
  nsplit = nsplit,
  importance = "permute",   # permutation-based VI
  ncores = ncores, verbose = FALSE
)

# VI vector lives in fit$VarImp
str(head(fit_perm$VarImp))
```

```{r perm-viz}
# Simple visualization
barplot(
  as.vector(fit_perm$VarImp),
  main = "Permutation Variable Importance (RLT)",
  ylab = "Importance",
  xlab = "Predictors",
  las = 2
)
```

## Option B - Distributed Assignment Importance

This configuration assigns importance using distributed attribution with OOB tracking. Use `importance = "distribute"` for distributed assignment importance.
```{r dist-importance}
fit_dist <- RLT(
  trainX, trainY, model = "regression",
  ntrees = ntrees, mtry = mtry, nmin = nmin,
  split.gen = rule, nsplit = nsplit,
  resample.prob = (trainn - 1)/trainn,   # leave-one-out style OOB
  resample.replace = FALSE,               # without replacement
  importance = "distribute",              # distributed assignment VI
  ncores = ncores, verbose = FALSE,
  param.control = list("resample.track" = TRUE)
)

# VI vector lives in fit$VarImp
str(head(fit_dist$VarImp))
```

```{r dist-viz}
barplot(
  as.vector(fit_dist$VarImp),
  main = "Distributed Variable Importance (RLT)",
  ylab = "Importance",
  xlab = "Predictors",
  las = 2
)
```
